---
title: "Kakcus-Turján"
author: "Clemens Schmid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spit attribution for small scale excavations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
if (!requireNamespace("plotly", quietly = TRUE)) {
  stop("plotly needed for this vignette to build.",
    call. = FALSE)
}
```

First we need to load recexcavAAR and some external packages:

- **dplyr**: filter function
- **kriging**: simple surface modelling tool
- **magrittr**: introduces piping to R via %>% operator
- **plotly**: nice, interactive plots via javascript implementation

```{r, message=FALSE}
devtools::load_all()
library(dplyr)
library(kriging)
library(magrittr)
library(plotly)
```

Like in the first vignette `Ifri el Baroud` we start by setting up an artificial and pretty simple excavation trench with a depth of 1.4 meters, a length of 6 meters and a width of 8 meters. This one is not parallel to the main axis of our coordinate system. 

```{r}
edges <- data.frame(
  x = c(6.899, 10.658, 4.428, 0.669, 6.899, 10.658, 4.428, 0.669),
  y = c(19.292, 14.616, 9.597, 14.273, 19.292, 14.616, 9.597, 14.273),
  z = c(9.7, 9.7, 9.7, 9.7, 8.3, 8.3, 8.3, 8.3)
)
```

We can plot the corner points of this trench, but first of all, we have to calculate a reasonable aspectratio. Remember: The trench is tilted in relation to the main axis. 

```{r fig.width=7, fig.height=5}
rangex <- abs(max(edges$x) - min(edges$x))
rangey <- abs(max(edges$y) - min(edges$y))

vis <- plot_ly(
  edges, x = x, y = y, z = z, 
  type = "scatter3d", mode = "markers",
  marker = list(size = 2, color = "blue", symbol = 104)
) %>% 
  layout(
    showlegend = FALSE,
    autorange = F, 
    aspectmode = 'manual', 
    scene = list(
      dragmode = "orbit",
      aspectratio = list(x = rangex, y = rangey, z = 5),
      camera = list(
        eye = list(x = -7, y = 7, z = 9) 
      )
    )
  ) 

vis
```

The approach for the excavation of this trench is to go deeper in squares of 1 meter by 1 meter. The spit depth is about 30 centimeters (rather unrealistic ;-)), but as we also try to follow natural layers, the individual spit depth varies a lot. Fortunately we have niveau measurements of the resulting surfaces. 

```{r}
KT_squarecorners
KT_spits
```