---
title: "recexcavAAR: Vignette >>Transforming Coordinates<<"
author: "Benjamin Serbe"
date: "Januar 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette gives an overview on the topic of 2D-Coordinate Transforming with the in recexcevARR implemented *cootrans-*function.


##Background

In archaeology you often have to deal with data acquired in an era before DGPS and tacheometry became the common surveying method. Even today some excavations use the traditional methods. These data are mostly recorded in local defined grids which are often not suitable for mapping in Geographic Information Systems.  
The *cootrans-*function provides a method to transform a set of local acquired coordinates into projected ones under given circumstances. 


##Requirements

This transformation method works only if you have at least two pairs of coordinates of which you know the local and the projected coordinates. Note that these coordinates have to be projected not geographic^1^. 


##Limitations

* This transformation method provides a two-dimensional transformation so only X- and Y-values will be calculated.
* This method is **not** suitable for datum transformations (e.g. converting from UTM to national grids).


##Example: A Simple transformation

First load the recexcavARR package:

```{r, message=FALSE}
library(recexcavAAR)
```

For this example we use a simple dataset. At first we have the data of the initial surveying with the three major measuring points placed by the surveyor. These are our **projected coordinates** (in this example UTM-coordinates).

```{r}
proj <- data.frame(
	E = c(578171.033, 578168.680, 578178.780),
	N = c(6017854.028, 6017859.595, 6017856.277)
)
```

After we constructed our own local grid, we add the corresponding local coordinates to the projected ones.

```{r}
coord <- data.frame(
	proj,
	X = c(1, 1, 9),
	Y = c(7, 1, 0)
)
```
This is the data table we need to successfully perform a transformation for all local coordinates we acquire throughout the excavation. 

Now we add some data e.g. the corners of our excavation trench or the artifacts we excavated. 

```{r}
measured_points <- data.frame(
	x = c(1,6,1,6),
	Y = c(3,3,5,5),
	point = c("corner1", "corner2", "corner3", "corner4")
)
```

And now let's use the *cootrans-*function to calculate the UTM-coordinates of our trench corners.

```{r}
absolute_data <- cootrans(coord, c(3,4,1,2), measured_points, c(1,2))
```
The first parameter in this function is our data.frame for the transformation. The directional shifts, rotation arcs and the scale will be calculated from these information.  
The second parameter is a vector of the column indices for our transform table in the specific order: 

1. the local x-value (east-value)

2. the local y-value (north-value)

3. the projected x-value (east-value)

4. the projected y-value (north-value)  

The third parameter is the data.frame with our measured local coordinates.  
Our last parameter is a vector of the column indices of our dataset in order:  

1. the local x-value (east-value)

2. the local y-value (north-value)  



The *cootrans-*function returns the original data.frame with two additional columns with the calculated projected coordinates and information about the transforation


In this case we an addtional warning message of high deviations within the calculation. The most likely mistake is the incorrect assignement of local and corresponding projected coordinates. This does not have to be the reason for the triggered warning message, because the internal control mechanism is rather strict. But to check the case of incorrectly defined points we can use the additional parameter "checking".

```{r, message=FALSE}
absolute_data <- cootrans(coord, c(3,4,1,2), measured_points, c(1,2), checking = TRUE)
```

In this case the function will return a `FALSE`-value and display two plots in seperate windows. The first with the mapped local coordinates and the second with the mapped projected coordinates. Every survey point is labled with an index number of their row.  
In this example we spot that the local point indices 1 and 2 are not corresponding with the projected ones. So they are likely mixed up. In this case we simply rebuild the transformation data.frame:

```{r}
coord <- data.frame(
	proj,
	X = c(1, 1, 9),
	Y = c(1, 7, 0)
)
```
In this case we only interchanged the Y-value of the first and second coordinate. After running the *cootrans*-function for the second time
```{r}
absolute_data <- cootrans(coord, c(3,4,1,2), measured_points, c(1,2))
```
everything seems fine. No warning messages pops up. 

***
**Attention!**  
The warning message and checking parameter only work reasonable if there are **more than two** given coordinate pairs. In case of only two points and incorrect defined coordinates the algorithm will just turn all of the measured points by 180°. So always check your output!  
Note that a warning message is also triggered in case you convert a survey with the parameters *one local unit* is not *one projected unit*^2^.
***


***
^1^ For those who are not familiar with these terms: projected coordinates use a grid system such as UTM. Geographic coordinates are e.g. "lon-lat"-coordinates.  
^2^ In most projected coordinate systems *one projected unit* equals 1 metre